Finch: A Datastructure-Driven Array Programming Language

### 2024-03-20 main.tex
Update on Overleaf.

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. 
\saman{Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. 
Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Arrays are the highest-
performing datastructure with a long history of investment and innovation, from hardware support to compiler
technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.  \\
Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data. Second, the arrays and loops abstraction introduced in FORTRAN supports a lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics. In this paper, we attempt to truly bring computing on structured data to the same Sparse tensor algebra can be put on the same compiler transformation and code generation footing as dense tensor algebra and array codes.  }
\end{abstract}

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. 
\saman{Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. 
Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Arrays are the highest-
performing datastructure with a long history of investment and innovation, from hardware support to compiler
technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.  \\
Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data. Second, the arrays and loops abstraction introduced in FORTRAN supports a lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics. \\
 In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as
sparse tensors, databases, and lossless compression
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  }
\end{abstract}
